---
title: "Intermediate_Homework"
author: "Joseph R. Stinziano"
date: "13/02/2020"
output: html_document
---

#1. Create a  dataset with 30 observations each of 5 species, 
with three different randomly generated measured variables where the
random sampling comes from a normal distribution. HINT: use ?runif
and to find out the names for other distributions. How might you
use a randomly-generated normal dataset in your analyses? How do
you ensure that this example is reproducible?
```{r}

```

#2. Extract data observations from the randomly generated dataset
that are greater than the mean of variable 1, and less than two
standard deviations from the mean of variable 2. Is there more
than one way to do this?
```{r}

```

#3. Create a function that runs a linear model of variable 1 ~
variable 2 for each of the five species separately, and outputs
the model coefficients to a properly labelled dataframe. Run the
function for the dataframes from 1 and 2. How might you modify
the function to extract the r-square value? What about standard
errors of the parameter estimates? Add these capabilities to
your function.
```{r}

```

#4. Create a function that generates QAQC plots of the data,
assigning each plot to a separate element of a list. Examples
of QAQC include boxplots for each variable, ratios of variables
(for example in most trees, we expect stem diameter to be less
than height), etc. QAQC: quality assurance & quality control.
What kind of QAQC plots might you use for your own data?
```{r}

```

#5. Make a function that generates a publication-quality,
multi-panel figure that shows all of the QAQC plots in 4.
This function should automatically print the plot to the
working directory. HINT: try the gridExtra or patchwork
packages
```{r}

```

#6. Using the nonlinear curve-fitting example in the lesson,
create a function that fits a linear model and the nonlinear
function that returns the better-fitting function based on
the lowest Bayesian Information Criterion (BIC).
```{r}

```