---
title: "Using R - Data Import & Wrangling"
author: "Joseph R. Stinziano"
date: "02/03/2020"
output: html_document
---

#Data Import & Wrangling
There are many ways to import data into R, however the easiest
is if the data are saved as a .csv file. You can save data in
this format directly from spreadsheet programs such as Excel.
In general, data should be in a "long format", meaning fewer
columns and more rows. For example, instead of having a column
for the height for each species, you would want a column for
species name, and a column for height.

R is object-oriented. In effect, this means that R uses objects
to manipulate and create other objects within an environment.
We can create and change new objects in R, and run calculations
on entire vectors, dataframes, matrices, or single elements
within vectors. We will run through examples below.

The example data we will be using is randomly generated, and
looks at how candy consumption, meme generate rate, and reddit
time vary across different types of scientists when they are
given caffeine and/or internet access.

```{r}
#We can read in data from a .csv file like this
#Note there are other read functions, including
#read.delim if your data are delimited by a
#something other than a comma. Type in
#?read.delim for more information
data <- read.csv("example.csv",
                 stringsAsFactors = FALSE)
#Notice how the file name is in quotes, and the
#stringsAsFactors argument is set to false. This
#makes sure that things that are not factors are
#not read in as such.

#If the data are in the .xls or .xlsx format
#we can use the read_excel function from the
#package readxl
#Load the package
library(readxl)
data_2 <- read_excel("example_excel.xlsx")
#However, it is much better to have the data in
#a .csv format because sometimes the read_excel
#function has unexpected effects on the values
#it reads in.
```

Summarizing data
```{r}
#####
#Summarizing
#We can take a look at data structure
str(data)

#and the first few lines
head(data)

#If we want to look at or use a single variable
#from a dataframe, we have two ways:
#dataframe$variable
data$species

#dataframes[, "variable"]
data[, "species"]

#We can look at a quick summary of our dataframe
summary(data)

#as well as for individual variables
summary(data$meme_rate)

#We can check the type of variable using:
class(data$meme_rate)

#We can also convert variables according to the
#family of as. functions
#If treatment1 and treatment2 are supposed to
#be factors, then:
data$treatment1 <- as.factor(data$treatment1)
data$treatment2 <- as.factor(data$treatment2)

#We can also summarize data into a new dataframe
#using tools from the dplyr package
#Load the package
library(dplyr)

#Below, the pipeline operator, %>%, passes
#everything on its left side to the calculation
#on the right side
data_summary <- data %>%
  group_by(species, treatment1,
           treatment2) %>% 
  summarize(meme_rate_mean = mean(meme_rate),
            meme_rate_se = sd(meme_rate) /
              sqrt(n()),
            meme_rate_median = median(meme_rate))

#If we want to summarize all the variables all
#at once:
data_summary <- data %>%
  group_by(species, treatment1,
           treatment2) %>% 
  summarise_at(.vars = names(.)[c(5:7)],
               .funs = c(mean = "mean",
                         IQR = "IQR",
                         median = "median",
                         sd = "sd"),
               na.rm = TRUE)

#If you want to export your data to a new .csv
#file, use:
write.csv(data_summary, "example_summary.csv")
```

Subsetting data
```{r}
#####
#Subsetting
#We may be interested in looking at meme rates
#only for biologists, in which case we can subset
data[data$species == "Biologist", ]$meme_rate

#If we want a summary, we can run:
summary(data[data$species == "Biologist", ]$meme_rate)

#This is starting to get complicated, so we can
#assign the data for biologists to a new object
data_biologist <- data[data$species == "Biologist",]
summary(data_biologist$meme_rate)

#If we wanted to subset the data only to include
#very productive meme-producing biologists, we
#could subsample data_biologist for meme rates
#equal to or greater than the 3rd quartile
#Extract value for 3rd quartile
third_quart <- summary(data_biologist$meme_rate)[5]
#Note the use of [] to extract the 5th element
#of the summary. This notation can be used to
#extract an element of any vector

#Now let's subset
data_biologist_productive <- 
  data_biologist[data_biologist$meme_rate >= third_quart,]

#We could also create a subset according to
#more than one condition - let's say we want
#biologists and chemists. We will use the
#OR operator, |
data_biol_chem <-
  data[data$species == "Biologist" |
         data$species == "Chemist",]

#If we want chemists that eat fewer than 15
#pieces of candy per day, we will use the AND
#operator, &
data_chem_low <-
  data[data$species == "Chemist" &
         data$candy_consumption < 15, ]

#Notice how we use the OR operator when we
#have multiple subsets from the same variable
#but use the AND operator when subsetting
#according to multiple variables
```

Vector calculations
```{r}
#####
#Vector calculations
#We can calculate and add new variables to our
#dataframes. We may want to calculate the meme
#efficiency of our biologists as meme_rate
#divided by reddit_time
data_biologist$meme_efficiency <-
  data_biologist$meme_rate / data_biologist$reddit_time

#We could also calculate weekly reddit time:
data_biologist$reddit_weekly <-
  data_biologist$reddit_time * 7

#Notice how we can calculate new vectors by
#running calculations on multiple vectors
#at once
```