---
title: "Beginner Homework"
author: "Joseph R. Stinziano"
date: "10/02/2020"
output: html_document
---

#Homework
This homework may require you to consult different resources in R.
Some resources are built into R, for example by using ?function,
you will get the R helpfile for that function. However, if you do
not know which function to use, you may have to use a search
engine to find an answer.

1. Read in the light response data setting stringsAsFactors 
to TRUE and FALSE. What variables are affected by this argument?
What potential (if any) issues could arise from this? When setting
stringsAsFactors to TRUE, how do you solve the resulting issues?
Fix the issue using code.
```{r}

```

2. Create a new dataframe containing only Poplar ID 1A at a rate
of 50 using two different methods of subsetting. Are there any
differences between the two methods?
```{r}

```

3. Using the dataframe generated in 2, create a new variable,
A / gsw, and plot it against PARi. The units for A / gsw are
micromoles of CO2 per mol of H2O per meter squared per second,
while for PARi they are micromoles of photons per meter squared
per second. Add labels to the graph using the appropriate
symbols and subscripting.
```{r}

```

4. Using the dataframe from 1, rearrange the dataframe from a
long format to a wide format. Under what circumstances could a
wide format be valuable? What are the difficulties in using the
wide format in R?
```{r}

```

5. Using datafrom 2, log-transform variable A, then run a linear
regression to test for interactions between PARi and gsw. What
issue is there with log-transforming the variable? Why? Are the
assumptions of linear regression met? Why or why not? Assuming the
assumptions are met, what are the conclusions of the statistical
test? Which information would you report in the results section of
a manuscript?
```{r}

```

6. Export a publication-quality jpeg of a graph of A as a function
of PARi for Poplar.ID 2A at a Rate of 100 and another graph at a
Rate of 500. The output must be a single file with two panels, with
panels labelled "a)" and "b)". Each graph must have a white background
with a font size of 14 for all text elements. The units of A and PARi
must be micromoles per meter squared per second, using appropriate
symbols. Make the points red for panel a) and blue for panel b) with
a black outline and circular shape. The output should have a height
to width ratio of 2 : 1 with a resolution of 600 dpi.
```{r}

```

7. Export the dataframe from 2 into a .csv file.
```{r}

```

8. Below are broken lines of code. Find the errors and fix them.
What was the mistake in each line of code?
```{r}

```

9. What packages did you need to complete the homework? Which
functions did you use from each package?
```{r}

```