---
title: 'Using R: An Introduction to Data Wrangling'
author: "Joseph R. Stinziano"
date: "08/01/2020"
output: html_document
---

#Types of objects in R
R is an object-oriented programming language. There are 
different types of objects, including dataframes, lists,
vectors, matrices, arrays, etc. A lot of analyses in
biology can be conducted working mostly with dataframes,
lists and vectors. It is important to keep track of object
types in R, since how you work with each type of object
varies.

There are different classes of variables as well, including
doubles (integers and numeric), characters, factors, POSIX 
(useful for dates and times), etc.
```{r}
#Create a vector from 1 to 10
#Note that the colon : creates a sequence
a <- 1:10

#Check its class
class(a)

#Create a string
#Note the concanetate function c(), and the repeat function
#rep()
b <- c(rep("a", 5), rep("b", 5))

#Check its class
class(b)

#Next we can take vectors a and b to turn them into a
#dataframe. Note the as.data.frame() function and the
#cbind() function - cbind = column bind, to bind rows
#use rbind() for "row bind"
c <- as.data.frame(cbind(a, b), stringsAsFactors = FALSE)

#Check its class
class(c)

#Create summary
summary(c)
#Notice that "a" is a character, but should be a number
#Let's fix that
c$a <- as.numeric(c$a)
#Note that the "$" extracts a variable from a dataframe
#We can also use:
c[, "a"]

#Create a boxplot
boxplot(a ~ b, data = c)
#Notice the "~" - this is used for specifying formulas
#This will make more sense when we get to the stats

#What happens if we do not use as.data.frame?
c <- cbind(a, b)
class(c)
summary(c)
#Note that if we try to extract a variable from a matrix
#like we do for a dataframe, it doesn't work!
c$a

#Instead we extract the column like this:
c[, "a"]

#To extract a row, we can use:
c[1, ]
#Notice how one method of extracting information is more
#general than others? In some cases you may want to use
#a specific method (like "$") to force an error if the
#object types are incorrect!

#What about factors?
c <- as.data.frame(cbind(a, b), stringsAsFactors = FALSE)
c$b <- as.factor(c$b)
c$b
summary(c$b)
levels(c$b)
#Note that factors are stored as dummy numbers referring
#to a character level. To convert from a factor to a
#character, use:
c$b <- levels(c$b)[c$b]
```

#Importing data into R
Unless you are doing specialized analyses, most of your data
will be in a delimited text format, including tab-delimited,
and comma-separated values (csv). In some case you will need
to read in an Excel file - luckily there are functions for
that as well.

When reading in files, it is best to set stringsasFactors as
FALSE to avoid the issue of have to convert data from a factor
to a character.

R looks for files based on the working directory (i.e. file
folder). Before importing data, it is important to confirm
that the working directory is correct.
```{r}
#Get the working directory
getwd()

#This line is for an example of how to set working directory
WD <- getwd()

#You can set the working directory with:
setwd(WD)

#It is easiest to read data into R when it is in the .csv format
data <- read.csv("light_responses.csv", stringsAsFactors = FALSE)

#If your data are in a tab-limited format, use:
data <- read.delim("filename.extension", sep = "\t", 
                   stringsAsFactors = FALSE)


```

#Rearranging data in R
```{r}

```

#Data visualization in R
There are two major approaches to plotting in R, the base plot
function and ggplot. ggplot is by far easier to perfect than
base plot.
```{r}
#Install ggplot2
install.packages("ggplot2")

#Load the package
library(ggplot2)

#Let's read in some data
data <- read.csv("light_responses.csv", stringsAsFactors = FALSE)

#Let's look at the structure of the data file
str(data)

#Let's plot the data based on rate
ggplot(data = data, aes(x = PARi, y = A, colour = Rate)) +
  geom_point()

#That's a little too much data for the plot, so let's try something
#else
data <- split(data, data$Poplar.ID)

ggplot(data = data[[1]], aes(x = PARi, y = A, colour = Rate)) +
  geom_point()

#Now let's clean this up a bit to make it look nice
figure_1 <- ggplot(data = data[[1]], aes(x = PARi, y = A, colour = Rate)) +
  geom_point() +
  labs(x = expression("Irradiance ("*mu*mol~m^{-2}~s^{-1}*")"),
       y = expression("A ("*mu*mol~m^{-2}~s^{-1}*")")) +
  theme_bw() +
  theme(text = element_text(size = 14),
        legend.position = c(0.8, 0.25),
        legend.background = element_blank())

figure_1

#Now how do we get it out of R?
#Highlight and run all of these lines
#File will appear in working directory
jpeg("Figure 1.jpeg", height = 4, width = 4, res = 600, units = "in")
figure_1
dev.off()

#Look to see if it appeared
list.files()

#What if you're feeling lazy and want a quick check on your data?
#Use the plot() function:
plot(A ~ gsw, data[[1]])
#This is nice when you need a fast visual check during your analysis

#Okay, but what if I want to use ggplot but am not too confident?
#Try ggplotgui!
install.packages("ggplotgui")
library(ggplotgui)

#ggplotgui gives you an interface with:
ggplot_shiny(data[[1]])
```

#Basic statistics in R
This section will cover linear regression, t tests, basic ANOVA,
and assumption checking
```{r}
#Let's read in some data
data <- read.csv("light_responses.csv", stringsAsFactors = FALSE)

#We are going to see the impact of gsw on Ci
plot(gsw ~ Ci, data[data$Poplar.ID == "1A", ])
#Looks like there is some grouping going on, probably due to
#the rate at which the light response was obtained
model_1 <- aov(data = data,
               gsw ~ Ci * Rate)
#Check assumptions:
plot(model_1)
#Data are not normal, probably should not use regular ANOVA

#We can take a look at the output for the example
summary(model_1)

#What about if we take a regression approach instead?
```

